Supplementary material

### 2020-05-02 Manuscript/ms.tex
Tidying

Differential equation models provide a natural way to describe mechanistic processes, but that description must also account for sources of uncertainty \cite{Hotelling1927, Wikle2010}.
In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
The structure of the processes themselves can be uncertain \cite{Ellner1998}, and needs to be informed by available, often noisy, data.
Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).

@@

Several of the parameters in this model are assumed to be fixed and known (Table 1).
The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria were taken from the 2010 census.
To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) were also fixed at literature values.
Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground \cite{weather}.

The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).  
Where possible, we specified informative prior distributions on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).

\begin{table}[!ht]
\label{parameters}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the mean of the prior distribution. See Supplemental Material for a full description of all prior distributions.}
\begin{tabular}{lp{6cm}lll}
Parameter & Description & Prior mean & Posterior mean & Citation\\
\hline
$N$ & Human population size & 327801 & & \cite{vitpop} \\
$1/d$ & Human life-span & 76 years & &\cite{vitlong} \\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & & \cite{Scott2000}\\
$1/\rho_{v}(t)$ & Extrinsic incubation period & $\exp\left(7.9 - 0.21 T(t) \right)$ days & & \cite{Chan2012}\\
$V_{E0}$ & Initial exposed mosquitoes &  0 & &\\
$V_{I0}$ & Initial infectious mosquitoes & 0 & &\\
\hline
$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & 0.88 (0.7, 1.1) &  \cite{Brady2013} \\
$1/\rho$ & Latent period in host & 0.87 weeks  & 1.72 (1.2, 2.3) &  \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & 3.6 (3.2, 4.1) & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  114 (72, 160) & \cite{Reich2013}\\
$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & 0.013 (0.01, 0.02) &\\
$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & 0.0005 (0.0004, 0.0008) & \\
$S_0$ & Proportion initially susceptible & 0.4 & 0.42 (0.28, 0.57)& \cite{Cardoso2011a} \\
$E_0$ & Number initially exposed & $100$ & 148 (104, 196) & \\
$I_0$ & Number initially infectious & $60$ & 79 (45, 116) &\\
$r_0$ & Initial mosquito population growth rate & 0 & 0.008 (-0.01, 0.02) &\\
$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & -0.16 (-0.4, 0.08) &\\
$V_{N0}$ & Initial mosquito population size & $2N$ & $1.5N$ $(1.1N, 1.9N)$ &\\
$\phi_y$ & Reporting probability & 0.083 & 0.14  (0.1, 0.18) &\cite{Silva2016}\\
$\log(\phi_q)$ & Log per-trap mosquito capture rate & $-13$  & -13.2 (-13.6, -13)  &\\
$\eta_y$ & Overdispersion of case reports & $0^*$ & 0.12 (0.1, 0.13) &\\
$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & 0.14 (0.12, 0.16)&\\
\end{tabular}
\end{center}
\end{adjustwidth}
$^*$ indicates prior mode, rather than mean.
\end{table}

@@

Combining the data, process, and parameter models \cite{Berliner1996}, we can summarize the full hierarchical model as:
\begin{align}
y_t | \cdot & \sim \text{NegBin}(\phi_y (C(t) - C(t-1)), \eta_y),
\\
q_t | \cdot &\sim \text{NegBin}(V_{C}(t) - V_{C}(t-1), \eta_q)\\
(\mathbf{C}, \mathbf{V_C}) & = \mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})\\
\epsilon_{rt} & \sim \text{Normal}(0, \sigma^2_r)\\
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
\boldsymbol{\theta} & \sim [\boldsymbol{\theta}]
\end{align}
where $\boldsymbol{\theta}$ is a vector of all the model parameters and initial conditions, and $\mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})$ represents the (numeric) solution to the differential equation model (Eq. 1-14) as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
Sampling from the posterior distribution of the parameters in mechanistic models is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
However, the variable $\epsilon_t$ introduce flexibility to the mechanistic model that 
soaks up lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
Gradient-based methods like Hamiltonain Monte Carlo (HMC) can then more easily and efficiently traverse the posterior.
Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
We ran 3 chains with different starting values for 5,000 iterations each, discarding the first 2,500 as burn-in.
Convergence diagnostics and mixing were evaluated using the shinystan package \cite{shinystan}.
Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
Timesteps as small as $1/8$ of a day were explored and did not qualitatively change the modeled dynamics.
Code is available from \href{https://github.com/clint-leach/mosquito-recon}{https://github.com/clint-leach/mosquito-recon}.

@@

Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of mosquito control applied in each week of the first four years of the time series.
Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
For each week and each posterior sample, we simulated the dynamics resulting from a $5\%$ reduction in the mosquito population implemented at the beginning of that week (affecting susceptible, exposed, and infectious mosquitoes equally).
To capture the likely rapid rebound in mosquito abundance following a single pulse of control \cite{Focks1987, Burattini2008}, we also simulated a $5\%$ increase in mosquito birth rate in the following week to return mosquito abundance to its previous trajectory (without this, the $5\%$ reduction in abundance persists indefinitely).
We then compared the number of cases produced during each year in the control scenario to the same number in the uncontrolled scenario.
A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.

@@

\begin{figure}[!h]
\includegraphics[scale = 1]{figures/fig2.eps}
\caption{{\bf Mosquito mortality and temperature.}
Posterior median mosquito mortality rate as a function of weekly mean temperature (Celsius).
}
\label{mortality}
\end{figure}

The effect of a given mosquito control intervention (i.e. the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Figure \ref{control}).
This variation in the effect of control was tightly correlated with the estimated mosquito mortality, with a median posterior correlation between the two time series of 0.96.
As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (though the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load coming from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implemented in week 34 (late August), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
This broadly corresponds to the time of year when both dengue case reports and mosquito trap counts are low.

@@

The variation around the predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
Thus mosquito control was more effective at reducing disease burden in simulations with long average mosquito lifespans (i.e., low mosquito mortality rates) or low overall prevalence (i.e., fewer undetected cases).

@@

The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
%In the dengue system, these drivers include waning human immunity, the race between the extrinsic incubation period of the virus and mosquito lifespan, and the seasonal and stochastic variation in these rates.
Seasonality in particular is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
However, the observed seasonality in transmission likely emerges from the combined effects of multiple seasonally-varying components that may be driven by different environmental factors and oscillate in different phases (e.g., mosquito abundance seems to lag slightly behind temperature-driven variation in extrinsic incubation period, Figure \ref{timeseries}).
Integrating these seasonally-varying components into a synthetic measure of transmission potential (e.g., a temperature-dependent effective reproduction number, \cite{Codeco2018}), or more specifically, a measure of the transmission potential of mosquitoes, is difficult.

\textbf{Mosquito mortality modulates the transmission potential generated by other mosquito and epidemiological processes}
Here, we positioned the latent mosquito mortality as the link between mosquito abundance, the extrinsic incubation period, and human cases.
In this way, the estimated mosquito mortality rate serves as an index of transmission potential, opening or closing the mosquito life history bottleneck \cite{Smith2012} as necessary to fit to the case reports data.
The resulting seasonality in the estimated trajectory suggests that the seasonality in mosquito abundance and the extrinsic incubation period was not sufficient to capture the observed case reports.
In particular, when mosquitoes were relatively scarce and transmission limited, we estimated a  relatively low mosquito mortality, suggesting that long-lived mosquitoes are required to maintain observed levels of transmission through the off-season.  
On the other hand, when mosquitoes were abundant, we estimated relatively high mosquito mortality rates, suggesting that transmission needed to be damped.

The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state and then bite a susceptible human.
In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
The tight correlation between our estimated mosquito mortality and the effect of control fits with this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
Specifically, our simulated mosquito control interventions were most effective at reducing the disease burden when applied around week 34 (i.e., early September), in the dengue off-season.
The effectiveness of this control was likely driven by the fact that transmission during the off-season was already limited by relatively few mosquitoes and a relatively high extrinsic incubation period.
Given that limitation, transmission was maintained by relatively few long-lived mosquitoes, making the system vulnerable to perturbation.

On the other hand, we found that a single pulse of control was relatively less effective when implemented during an outbreak, when mosquitoes were abundant (and unlikely to be limiting transmission) but short-lived.
Given the relatively high mosquito mortality rates during this time, exposed and infectious mosquitoes were already fairly ephemeral, such that the relatively small disruption induced by control likely made little difference.
Moreover, due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population, the population of exposed mosquitoes likely recovered quickly \cite{Newton1992, Burattini2008}.
Barsante \emph{et al.} \cite{Barsante2015} and Oki \emph{et al.} \cite{Oki2011} similarly found that control was most effective when applied well before peak prevalence, either during the dry season \cite{Barsante2015} or early in the rainy season \cite{Oki2011} (September is near the end of the dry season in Vit\'oria).
While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.

@@

The posterior correlation between the effectiveness of control, the case reporting rate, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity (and thus fewer susceptibles) generated systems in which transmission was more difficult to disrupt with control actions.
As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend on our ability to monitor and predict the dynamics of human immunity.
To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016} and the number of unreported cases \cite{Silva2016}..
In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR (Time-series Susceptiple-Infected-Recovered) frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.

@@

The periodic structure in the time series of the posterior means of the $\epsilon_{\nu}$ (Supplemental Figure 1) suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values). 
The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control intervention), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different risk levels).

@@

In addition to epidemiological complexity, dengue dynamics are further complicated by the reporting process.
We estimated a relatively long average intrinsic incubation period ($1/\rho$, posterior mean of 1.7 weeks) relative to our prior mean (0.87 weeks), suggesting possible reporting delays \cite{Reich2016}.
The case data to which we fit the model represent reports of ``dengue-like illness,'' without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g., chikungunya or Zika).
However, neither chikungunya nor Zika had emerged as substantial public health threats in Brazil by the end of our time series in late 2012 \cite{PAHO2014, PAHO2015}.
In addition, given the high underreporting rate expected for dengue fever \cite{Silva2016}, and the uncertainty incorporated into the measurement model, we expect misreported cases to have a small effect on our analyses.

Mosquito control interventions can prevent cases by acting on any of the components of vectorial capacity.
We focused on the direct effect of killing adult mosquitoes on transmission, but adult control can also act by reducing egg laying and the number of mosquitoes in the next generation \cite{Brady2016}.
Given the relative simplicity of our mosquito model, we were unable to eplore the feedbacks that adult control may induce in mosquito population dynamics, and we instead assume that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
We expect that capturing these feedbacks would likely reinforce our conclusions about the utility of off-season control, as the disruption to mosquito population dynamics would more strongly limit the ability of the mosquito population to maintain transmission through the off-season.

>>>>

Differential equation models provide a natural way to describe mechanistic processes, but that description must also account for sources of uncertainty \cite{Hotelling1927, Wikle2010}.
In particular, the values of parameters (e.g., the average length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
The structure of the processes themselves can be uncertain \cite{Ellner1998}, and needs to be informed by available, often noisy, data.
Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).

@@

Several of the parameters in this model are assumed to be fixed and known (Table 1).
The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria were taken from the 2010 census.
To maintain identifiability, the transmission rate ($\lambda$) was also fixed at literature values.
Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground \cite{weather}.

The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_y$, $\phi_q$, $\eta_y$, $\eta_q$).  
Where possible, we specified informative prior distributions on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).

\begin{table}[!ht]
\label{parameters}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random.  The values in parentheses after the posterior means give the 80$\%$ credible interval. See Supplemental Material for a full description of all prior distributions.}
\begin{tabular}{lp{6cm}lll}
Parameter & Description & Prior mean & Posterior mean & Citation\\
\hline
$N$ & Human population size & 327801 & & \cite{vitpop} \\
$1/d$ & Human life-span & 76 years & &\cite{vitlong} \\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & & \cite{Scott2000}\\
$1/\rho_{v}(t)$ & Extrinsic incubation period & $\exp\left(7.9 - 0.21 T(t) \right)$ days & & \cite{Chan2012}\\
$V_{E0}$ & Initial exposed mosquitoes &  0 & &\\
$V_{I0}$ & Initial infectious mosquitoes & 0 & &\\
\hline
$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & 0.88 (0.7, 1.1) &  \cite{Brady2013} \\
$1/\rho$ & Latent period in host & 0.87 weeks  & 1.72 (1.2, 2.3) &  \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & 3.6 (3.2, 4.1) & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  114 (72, 160) & \cite{Reich2013}\\
$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & 0.013 (0.01, 0.02) &\\
$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & 0.0005 (0.0004, 0.0008) & \\
$S_0$ & Proportion initially susceptible & 0.4 & 0.42 (0.28, 0.57)& \cite{Cardoso2011a} \\
$E_0$ & Number initially exposed & $100$ & 148 (104, 196) & \\
$I_0$ & Number initially infectious & $60$ & 79 (45, 116) &\\
$r_0$ & Initial mosquito population growth rate & 0 & 0.008 (-0.01, 0.02) &\\
$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & -0.16 (-0.4, 0.08) &\\
$V_{N0}$ & Initial mosquito population size & $2N$ & $1.5N$ $(1.1N, 1.9N)$ &\\
$\phi_y$ & Reporting probability & 0.083 & 0.14  (0.1, 0.18) &\cite{Silva2016}\\
$\log(\phi_q)$ & Log per-trap mosquito capture rate & $-13$  & -13.2 (-13.6, -13)  &\\
$\eta_y$ & Overdispersion of case reports & $0^*$ & 0.12 (0.1, 0.13) &\\
$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & 0.14 (0.12, 0.16)&\\
\end{tabular}
\end{center}
\end{adjustwidth}
$^*$ indicates prior mode, rather than mean.
\end{table}

@@

Combining the data, process, and parameter models \cite{Berliner1996}, we can summarize the full hierarchical model as:
\begin{align}
y_t | \cdot & \sim \text{NegBin}(\phi_y (C(t) - C(t-1)), \eta_y),
\\
q_t | \cdot &\sim \text{NegBin}(V_{C}(t) - V_{C}(t-1), \eta_q)\\
(\mathbf{C}, \mathbf{V_C}) & = \mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})\\
\epsilon_{rt} & \sim \text{Normal}(0, \sigma^2_r)\\
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
\boldsymbol{\theta} & \sim [\boldsymbol{\theta}]
\end{align}
where $\boldsymbol{\theta}$ is a vector of all the model parameters and initial conditions, and $\mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})$ represents the (numeric) solution to the differential equation model (Eq. 1-14) as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
Sampling from the posterior distribution of the parameters in mechanistic models is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
However, the variable $\epsilon_t$ introduce flexibility to the mechanistic model that 
soaks up lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
Gradient-based methods like Hamiltonain Monte Carlo (HMC) can then more easily and efficiently traverse the posterior.
Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
We ran 3 chains with different starting values for 4,000 iterations each, discarding the first 2,000 as burn-in.
Convergence diagnostics and mixing were evaluated using the shinystan package \cite{shinystan}.
Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
Timesteps as small as $1/8$ of a day were explored and did not qualitatively change the modeled dynamics.
Code is available from \href{https://github.com/clint-leach/mosquito-recon}{https://github.com/clint-leach/mosquito-recon}.

@@

Given a subset of the samples from the posterior distribution as obtained above (2000, taken to reduce computation time), we simulated the effects of a single pulse of mosquito control applied in each week of the first three years of the time series.
Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden over the following year.
For each week and each posterior sample, we simulated the dynamics resulting from a $5\%$ reduction in the mosquito population implemented at the beginning of that week (affecting susceptible, exposed, and infectious mosquitoes equally).
To capture the likely rapid rebound in mosquito abundance following a single pulse of control \cite{Focks1987, Burattini2008}, we also simulated a $5\%$ increase in mosquito birth rate in the following week to return mosquito abundance to its previous trajectory (without this, the $5\%$ reduction in abundance persists indefinitely).
We then compared the number of cases produced over the year following the control intervention in the control scenario to the same number in the uncontrolled scenario.
A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.

@@

\begin{figure}[!h]
\includegraphics[scale = 1]{figures/fig2.eps}
\caption{{\bf Mosquito mortality and temperature.}
Posterior median mosquito mortality rate as a function of weekly mean temperature (degrees Celsius).
}
\label{mortality}
\end{figure}

The effect of a given mosquito control intervention (i.e. the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Figure \ref{control}).
This variation in the effect of control was tightly correlated with the estimated mosquito mortality, with a median posterior correlation between the two time series of 0.96.
As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (though the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load coming from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implementedaround  week 34 (late August/early September), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
This broadly corresponds to the end of the dry season in Vit\'oria, when both dengue case reports and mosquito trap counts are low.

@@

The variation around the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
Thus mosquito control was more effective at reducing disease burden in simulations with long average mosquito lifespans (i.e., low mosquito mortality rates) or low overall prevalence (i.e., fewer undetected cases).

@@

The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
Seasonality in particular is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
However, the observed seasonality in transmission likely emerges from the combined effects of multiple seasonally-varying components that may be driven by different environmental factors and oscillate in different phases (e.g., mosquito abundance seems to lag slightly behind temperature-driven variation in extrinsic incubation period, Figure \ref{timeseries}).
Integrating these seasonally-varying components into a synthetic measure of transmission potential (e.g., a temperature-dependent effective reproduction number, \cite{Codeco2018}), or more specifically, a measure of the transmission potential of mosquitoes, is difficult.

Here, we positioned the latent mosquito mortality as the link between mosquito abundance, the extrinsic incubation period, and human cases.
In this way, the estimated mosquito mortality rate serves as an index of transmission potential, opening or closing the mosquito life history bottleneck \cite{Smith2012} as necessary to fit to the case reports data.
The resulting seasonality in the estimated trajectory suggests that the seasonality in mosquito abundance and the extrinsic incubation period was not sufficient to capture the observed case reports.
In particular, when mosquitoes were relatively scarce and transmission limited, we estimated a  relatively low mosquito mortality, suggesting that long-lived mosquitoes were required to maintain observed levels of transmission through the off-season.  
On the other hand, when mosquitoes were abundant, we estimated relatively high mosquito mortality rates, suggesting that transmission needed to be damped.

The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state and then bite a susceptible human.
In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
The tight correlation between our estimated mosquito mortality and the effect of control fits with this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
Specifically, our simulated mosquito control interventions were most effective at reducing the disease burden when applied around week 34 (i.e., early September), in the dengue off-season.
The effectiveness of this control was likely driven by the fact that transmission during the off-season was already limited by low mosquito abundance and a relatively high extrinsic incubation period.
Given that limitation, transmission was maintained by relatively few long-lived mosquitoes, making the system vulnerable to perturbation.

On the other hand, we found that a single pulse of control was relatively less effective when implemented during an outbreak, when mosquitoes were abundant (and unlikely to be limiting transmission) but short-lived.
Given the relatively high mosquito mortality rates during this time, exposed and infectious mosquitoes were already fairly ephemeral, such that the relatively small disruption induced by control likely made little difference.
Moreover, due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population, the population of exposed mosquitoes likely recovered quickly \cite{Newton1992, Burattini2008}.
Barsante \emph{et al.} \cite{Barsante2015} and Oki \emph{et al.} \cite{Oki2011} similarly found that control was most effective when applied well before peak prevalence, either during the dry season \cite{Barsante2015} or early in the rainy season \cite{Oki2011} (September is near the end of the dry season in Vit\'oria).
While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
Further, although large pulses of imported cases could potentially swamp the effects of early control, in additional simulations we found that our results were robust to the import 10 infectious humans just before the annual outbreak.

@@

The posterior correlation between the effectiveness of control, the case reporting rate, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity (such that susceptibles replenished more slowly) generated systems in which transmission was more difficult to disrupt with control actions.
As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend in part on our ability to monitor and predict the dynamics of human immunity.
To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016} and the number of unreported cases \cite{Silva2016}.
In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR (Time-series Susceptiple-Infected-Recovered) frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.

@@

The periodic structure in the time series of the posterior means of the $\epsilon_{\nu}$ (Supplemental Figure 1) suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values). 
The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control intervention), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different mixing or risk levels).

@@

In addition to epidemiological complexity, dengue dynamics are further complicated by the reporting process.
We estimated a relatively long average intrinsic incubation period ($1/\rho$, posterior mean of 1.7 weeks) relative to our prior mean (0.87 weeks), suggesting possible reporting delays \cite{Reich2016}.
Moreover, the case data to which we fit the model represent reports of ``dengue-like illness,'' without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g., chikungunya or Zika).
However, neither chikungunya nor Zika had emerged as substantial public health threats in Brazil by the end of our time series in late 2012 \cite{PAHO2014, PAHO2015}.
In addition, given the high underreporting rate expected for dengue fever \cite{Silva2016}, and the uncertainty incorporated into the measurement model, we expect misreported cases to have a small effect on our analyses.

Mosquito control interventions can prevent cases by acting on any of the components of vectorial capacity.
We focused on the direct effect of killing adult mosquitoes on transmission, but adult control can also act by reducing egg laying and the number of mosquitoes in the next generation \cite{Brady2016}.
Given the relative simplicity of our mosquito model, we were unable to eplore the feedbacks that adult control may induce in mosquito population dynamics, and instead assumed that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
We expect that capturing these feedbacks would likely reinforce our conclusions about the utility of off-season control, as the disruption to mosquito population dynamics would more strongly limit the ability of the mosquito population to maintain transmission through the off-season.
